Changes in the justification of educational inequalities. The role of perceptions of inequality and meritocracy during the COVID pandemic.

### 2025-04-17 submit/Title-Page.tex
update submit

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  12pt,
  10pt]{article}

\usepackage{amsmath,amssymb}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=0.78in]{geometry}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{times}
\usepackage{caption}
\captionsetup[figure, table]{labelfont={bf},labelformat={default},labelsep=period}
\usepackage[figurename=Fig.]{caption}
\usepackage{graphicx}
\usepackage{float}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage[noblocks]{authblk}
\renewcommand*{\Authsep}{, }
\renewcommand*{\Authand}{, }
\renewcommand*{\Authands}{, }
\renewcommand\Affilfont{\small}
\renewcommand\Affilfont{\small}
\usepackage{mathptmx} % Times
\usepackage{times}
\usepackage{caption}
\captionsetup[figure, table]{labelfont={bf},labelformat={default},labelsep=period}
\usepackage[figurename=Fig.]{caption}
\usepackage{graphicx}
\usepackage{float}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Changes in the justification of educational inequalities. The role of perceptions of inequality and meritocracy during the COVID pandemic. },
  pdfauthor={Juan Carlos Castillo; Julio Iturra; Kevin Carrasco},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Changes in the justification of educational inequalities. The
role of perceptions of inequality and meritocracy during the COVID
pandemic. \footnote{Corresponding author (juancastillov@uchile.cl -
  \href{https://juancarloscastillo.github.io/jc-castillo/}{jc-castillo.com}
  - Ignacio Carrera Pinto 1045 Ñuñoa, Santiago de Chile). The authors
  recognize the support of the National Research and Development Agency
  (ANID) through the grant from the National Fund for Scientific and
  Technological Development (FONDECYT) N° 1210847 and N° 1250518; from
  the Center for Conflict and Social Cohesion Studies (COES) grant
  Financing Fund for Research Centers in Priority Areas (FONDAP) N°
  1523A0005; and Millenium Nucleus on Digital Inequalities and
  Opportunities (NUDOS) NCS2022\_046}}

  \author{Juan Carlos Castillo}
            \affil{%
                  Universidad de Chile
              }
          \affil{%
                  Centro de estudios del conflicto y cohesión social
                  (COES)
              }
          \affil{%
                  Núcleo milenio de desigualdades y oportunidades
                  digitales (NUDOS)
              }
        \author{Julio Iturra}
            \affil{%
                  International Graduate School of Social Sciencies
                  (BIGSSS), University of Bremen, Germany
              }
        \author{Kevin Carrasco}
            \affil{%
                  Centro de estudios del conflicto y cohesión social
                  (COES)
              }

\date{}
\begin{document}
\maketitle

@@

\newpage

\section{Abstract}\label{abstract}

Education is considered a key tool for social mobility and equality of
opportunities. However, despite the widespread value of education,
disparities in educational outcomes persist over time. This paper is
guided by the following research questions: To what extent are such
educational disparities justified in society? Is the justification of
educational inequality affected in periods of personal and social
vulnerability, as in the COVID pandemic? And, what are the main factors
driving this kind of justification? The Chilean case offers an
interesting context for this study given its high economic inequality,
deep neoliberal policies, and commodification of social services such as
education. This economic and cultural environment has promoted
meritocratic ideals, whereby individual talent and effort are considered
key to get ahead in life, disregarding opportunities linked to social
origin. The central argument of this article is that the justification
of inequalities weakens during periods of vulnerability and crisis (such
as the health and economic crisis resulting from the COVID-19 pandemic).
Furthermore, we argue that such changes could be linked to a challenge
of meritocratic ideals. For testing the research hypotheses we estimate
a series of longitudinal multilevel models with data from a Chilean
longitudinal panel survey (2016 -- 2023, 6 waves, N = 2,927). Whereas we
find support for the association between the perception of meritocracy
and justification of inequality, the analyses that involved changes over
time showed an increase in the justification of inequality in education.
The discussion of these results delves into the social consequences of
justifying inequality in a sensitive area as education, as well as the
persistence of meritocratic ideals despite challenging events.\newline
\textbf{Keywords}: meritocracy, social inequality, inequality
justification, COVID-19

\section{Introduction}\label{introduction}

Despite the widespread recognition of education as a key factor for
social mobility and equality of opportunities, disparities in this
domain are far from being eliminated. As the last PISA report points
out, ``Socio-economically disadvantaged students are seven times more
likely than advantaged students to score below Level 2 in mathematics on
average across OECD countries'' (OECD 2023). Furthermore, empirical
evidence from social science research has revealed an exacerbation of
educational inequalities across various domains, such as family
socioeconomic status, gender, ethnicity, and migratory background,
particularly after COVID-19 (Darmody, Smyth, and Russell 2021).
Nevertheless, and despite the consistent empirical evidence on
educational inequalities and their consequences, research on
distributive justice indicates remarkable individual differences in the
justification of unequal access to education (Lee and Stacey 2023). In
the present paper, we argue that the perception of meritocracy in
society might play a role in the justification of economic inequalities
in this particular domain. We examine this association in the context of
COVID 19 in Chile, a period characterized by several challenges
regarding the political and economic system that could have affected the
way in which the role of individual performance and social outcomes are
deemed.

The Chilean case is particularly relevant when it comes to the study of
inequalities in education. The country is characterized by a high level
of economic inequality associated with the privatization and
commodification of several social policy domains, among them the
educational system (Bellei 2013; Corvalán, Carrasco, and García-Huidobro
2016). The neoliberal reforms implemented during and since the military
dictatorship (1973-1989) have deepened educational policies based on
private subsidies, achievement incentives, selection, and segregation
(Joiko 2019). Previous research in Chile has emphasized the concept of
an ``educational market'' (Corvalán, Carrasco, and García-Huidobro
2016), characterized by selection, competition, and the provision of
goods and services associated with payment capacities. Many studies have
addressed this problem from economic, political, and sociological
perspectives, generating relevant evidence on their academic and social
consequences (Ramos 2022). Likewise, massive social mobilizations in the
country (particularly in 2006, 2011, and 2018) have had educational
inequalities as one of the main targets, putting pressure on governments
towards a series of reforms that have advanced free higher education
reforms and have diminished the role of private actors in the
educational system (Donoso 2016; Reyes-Housholder and Roque 2019).

This research endeavors to dissect the nuanced alterations in the
justifications for educational disparities within the Chilean framework.
Central to our investigation is the premise that the rationale
underpinning these inequalities experiences a significant recalibration
during periods of vulnerability and crises, specifically under the
unprecedented strains of the global health and economic cataclysm
triggered by the COVID-19 pandemic (Hankivsky, Morrow, and Varcoe 2022;
Breznau 2021). A pertinent aspect of this discourse is articulated in
the works of Mijs (2016, 2018), showing the interplay between
meritocratic beliefs and the legitimization of disparities, as well as
the potential erosion of meritocratic principles amidst crises. This
line of research has highlighted how meritocratic beliefs are associated
with justifying unequal access to resources and rewards in society, by
legitimizing social hierarchies in which those who already have
resources and advantages are more likely to succeed (Sandel 2020;
McNamee and Miller 2004; Hadjar 2008). Within this framework, we point
out that periods of vulnerability (such as the health and economic
crisis resulting from the COVID-19 pandemic) could weaken the
justification of social inequalities, a change that would be at least in
part linked to a challenge of the meritocratic ideals.

\section{The justification of educational
inequalities}\label{the-justification-of-educational-inequalities}

The justification of economic inequalities is a controversial field of
study. As equality is usually considered a synonym of justice in common
sense, the idea that someone could be willing to justify inequalities
seems (in principle) far from what is considerably reasonable.
Nevertheless, and as pointed out by Barry, the whole normative
discussion in social justice can be conceived as ``the defensibility of
unequal relations between people'' (1989, 3), whereby empirical research
has consistently shown strong variations between individuals and
societies regarding inequality justification (Trump 2018; Kelley and
Evans 2021; Mijs 2021). Both theoretical and empirical approaches have
been able to establish a clear distinction between inequality and
justice, opening the question of to what extent inequality can be
tolerated or even justified.

Studies addressing inequality justification have assessed the support
for market ideologies, salary gaps, and distributive preferences. A
strand of this research has focused on specific policy areas such as
health, pensions, and education, attending to the degree of
justification of market-based access to social services. For instance,
Lindh (2015) found that public support for the market distribution of
social services is generally low in most countries, implying that most
people believe it is unfair for market forces to determine access to
basic social services. Similarly, Busemeyer and Iversen (2020) shows
that high-income citizens are less supportive of expanding public
spending on universal benefit schemes. Instead, they tend to prefer a
public basic insurance scheme, even if it provides relatively more
benefits to low-income individuals.

In the area of survey studies, the operationalization of educational
inequality justification in terms of questionnaire items has been
diverse, and there is not a single instrument or scale dedicated
specifically to this concept. Rather, some attitudinal surveys usually
include education-related items that are posteriorly used as dependent
or independent variables. Some examples are the International Social
Survey Programme (ISSP) (\emph{Is it just or unjust -- right or wrong --
that people with higher incomes can buy better education for their
children than people with lower incomes?}), and the European Social
Survey (ESS) 2018 (\emph{Compared to other people in {[}country{]}, I
have had a fair chance of achieving the level of education I was
seeking}). Although both questions have been used to measure the
justification of educational inequality, they refer to different
aspects: the ESS can be classified as reflexive, as the respondent is
involved as a referent for the question, whereas the ISSP question is
non-reflexive. Furthermore, ISSP refers to educational access, while the
ESS refers to educational outputs. Although such items covering
inequality justification are present in emblematic comparative studies,
still there are few studies focusing specifically in the justification
of educational inequality.

\emph{The COVID pandemic and public concerns about inequality}

The time frame for this study covers between 2016 and 2023, by using
longitudinal panel survey data provided by the ELSOC database (Estudio
Longitudinal Social de Chile). Although this period comprises the COVID
pandemic, the changes in the variables analyzed are not possible to be
attributed only to this event in terms of a health crisis, as several
economic, social and cultural processes overlapped worldwide. However,
and as several studies have shown, this period appeared to trigger
attitudinal changes that are relevant to explore, both in themselves and
as antecedent to future studies. In this sense, in this paper we argue
that the increase of risk perceptions in critical situations (such as
the pandemic) impacts public concerns about equality and redistribution
(Breznau 2021).

In Chile, the COVID-19 pandemic posed a significant threat to
economically vulnerable groups and those in precarious labor market
positions (Ministerio de Desarrollo Social y Familia 2023; Instituto
Nacional de Estadísticas 2022). The poverty rate, which was 11.2\% in
2015, decreased to 8.5\% by 2017, with a 10/10 ratio of 31.6. However,
during the pandemic in 2020, poverty rose to 10.7\%, the 10/10 ratio
spiked to 301.7, and unemployment reached a decade-high of 10.5\%. By
2022, poverty declined to 6.5\%, the 10/10 ratio improved to 61.2, and
unemployment decreased to 7.9\%, with a slight rise to 8.7\% in 2023.

Chile implemented sanitary controls early in the pandemic, initiating
lockdowns on March 18, 2020, and lifting all restrictions by March 5,
2023. In this context, Chile stands out as one of the strictest
countries globally and within South America in terms of pandemic-related
policies. According to the Stringency Index (Hale et al. 2021)---ranging
from 0 to 100 (100 being the strictest) and encompassing measures like
school closures, workplace shutdowns, and travel bans---the global
median was 42.54, while Chile scored 58.1. This placed Chile 36th among
184 countries and 4th in South America, following Peru (60.1), Suriname
(60.5), and Venezuela (80.5). Besides, an important cycle of protests
known as the ``social outburst'' took place between October 2019 and
March 2020 (Somma et al. 2021). As a consequence there were two
constitutional processes happening in this very same period after the
massive social upheavals. However, the protest drastically decreased as
a result of the lockdown (Joignant et al. 2020).

In terms of public views on economic inequality, it has been documented
that critical views toward inequality increased in 2019 and deepened
under the pandemic (COES 2023), which we argue can be in part explained
by the increased risk exposure and vulnerability in the context of the
COVID-19 pandemic. Although the pandemic might have triggered such
concerns mostly regarding health, we believe that other policy areas,
such as education, could have been affected in this period.
Consistently, our first hypothesis is:

\(H_1\): The justification of educational inequalities decreases after
the pandemic

Besides proposing a general decrease over time in the justification of
inequalities, a second hypothesis refers to specific aspects of the
pandemic. As some political measures from the state included several
social benefits (as delivering food boxes, free vaccination, direct
transfers, among others), we argue that being recipient of such benefits
could affect attitudes towards the redistributive role of the state as
equalizer opportunities to face risks. As a consequence, preferences for
educational inequalities would be reduced:

\(H_2\): Recipient of state benefits show lower justification of
educational inequalities

In the same line, we expect that those who give priority to take care of
the health of the population instead of the country economy would
exhibit larger preferences for educational inequalities:

\(H_3\): Those prioritizing health instead of the economic situation
show higher justification of educational inequalities

\emph{Meritocracy and educational inequalities}

Whether inequalities in education are considered just or not lead us to
take into account the concepts of merit and meritocracy. Meritocracy, a
concept coined by the sociologist Michael Young in his satirical work
``The Rise of the Meritocracy'' (1962), has a rich history and
multifaceted meaning. Young's term was initially used to criticize a
hypothetical future society where social and economic positions were
entirely determined by individuals' merits, talents, and efforts,
ostensibly eliminating the influence of inherited privilege. However,
over time, the term has evolved to describe a social system where
advancement and success are perceived as directly linked to one's
abilities and achievements, sidelining inequalities of opportunities
(Bell 2020). This would contribute to the justification of inequality in
the educational domain (Sandel 2020), as individual ability is often
considered inherited, and educational attainment is influenced by family
background (Goldin 2000).

Although there are few studies relating meritocratic beliefs and
inequality justification, from a broader perspective it is possible to
find research which incorporates the support of different distributive
principles with regard to education. For instance, in an experiment by
Igliozzi, Granot, and Ottati (2024), they establish how the support for
different justice principles (equity, equality, and need) is related to
support for redistributive policies in domains such as education. They
find that a larger system justification (Jost and Hunyady 2003) is
linked to higher support for the principles of equity and equality (over
need) in distributing educational outcomes, where equity was measured in
terms of access to education related to performance (`Students who place
within the highest 50\% of testing scores from the previous year can
attend the magnet school'). In the same line, Lee and Stacey (2023) show
that performance assessments (in what they call ``neoliberal
orientations'') predicted perceived fairness of educational inequality.

More specific empirical research examining the relationship between
meritocratic perceptions and the justification of educational inequality
provides valuable insights into the dynamics shaping individuals'
attitudes and beliefs. Some studies have explored the extent to which
people believe that educational systems are meritocratic and how these
perceptions influence their acceptance or rejection of educational
disparities, as well as status differences (McNamee and Miller 2004).
This belief is particularly influential in educational institutions,
where it not only encourages academic success but also legitimizes
social and income inequality (Batruch et al. 2023). Most of the evidence
so far suggests that those who believe in a direct link between effort,
talent, and educational success are more inclined to rationalize
existing disparities in educational outcomes. However, empirical
findings also reveal nuances in this relationship. For instance, studies
show that when individuals perceive structural barriers, such as
economic disparities or discrimination, as hindering the meritocratic
ideal, they may be less likely to justify educational inequalities (Day
and Norton 2023). This suggests that while meritocratic beliefs play a
significant role, individuals' awareness of broader societal issues can
mediate their willingness to rationalize educational disparities.

Empirical cross-sectional evidence from Chile reveals two significant
trends: low perceived meritocracy and limited justification for unequal
access to social services such as education and health. For instance,
Castillo, Madero-Cabib, and Salamovich (2013) analyzed data from the
International Social Survey Programme (ISSP) in 1999 and 2009,
highlighting consistently low support for inequality based on resource
allocation or individual merit. Similarly, Castillo, Miranda, and
Carrasco (2012) demonstrated widespread skepticism among Chileans
regarding the meritocratic system, with many disagreeing with the notion
that rewards are fairly distributed based on talent and effort. In line
with this evidence, we propose that:

\(H_4\): Higher meritocratic perceptions are positively related to the
justification of educational inequalities.

Combining the arguments regarding the effect of the pandemic and its
relationship with meritocracy, we explore interactions between
meritocracy and inequality justification over time. As COVID-19 pandemic
has affected the structural opportunities to attend schools,
universities, or workplaces, the ideal of meritocracy might have been
challenged in the time frame analyzed. As a consequence, the
justification of inequality could show a decrease over time:

\(H_5\): The positive association between meritocracy and inequality
justification mitigates along time

Along with meritocratic perceptions, in this study we incorporate
inequality perception. As pointed out by Janmaat (2013), justifications
are usually confused under the labels of ``beliefs'' and ``attitudes,''
and their distinction from perceptions is relevant as it helps state
that justifying a certain level of inequality requires taking into
account which level of (perceived) inequality is first and foremost
being recognized (Son Hing et al. 2011; Castillo et al. 2023). For
instance, a recent study by Day and Norton (2023) on the perception and
justification of inequality in university endowments in the U.S. points
out that individuals tend to underestimate the magnitude of existing
inequality while simultaneously desiring greater equality. Furthermore,
through an experiment, they demonstrate that information about
inequality in university endowments (i.e., manipulating inequality
perception) increases the perception of injustice in education. Results
of this study are in line with the consistent evidence on biases related
to the underestimation of economic inequality (Castillo, García-Castro,
and Venegas 2022; Gimpelson and Treisman 2018). As a larger inequality
in this domain threatens a basic equality ideal, the corresponding
hypothesis in this case is:

\(H_6\): A larger inequality perception is negatively associated with
the justification of educational inequalities.

Next, we take into account the evidence that relates rational interests
to justifications and preferences (Son Hing et al. 2019). Research about
the justification of educational inequality usually focuses on
differences between status groups, based on the rational premise that
those who are worst-off would tend to show lower levels of justification
(Busemeyer and Iversen 2020). For instance, Valant and Newark (2016)
find that Americans are more concerned about --- and more supportive of
proposals to close --- wealth-based achievement gaps than Black-White or
Hispanic-White gaps. In line with this evidence, we expect that:

\(H_7\): Those with lower status characteristics show lower
justification of educational inequalities

\section{Data, variables \& methods}\label{data-variables-methods}

\subsection{Data}\label{data}

The main data source is the Chilean Longitudinal Social Survey ELSOC
(2016--2023). ELSOC has been designed as a yearly panel study to
evaluate how individuals think, feel, and behave regarding a set of
social issues concerning conflict and social cohesion in Chile. The
sampling design is probabilistic, stratified, clustered, and multistage.
It provides adequate coverage of the country's largest cities
(Metropolitan Area of Santiago, Valparaíso, and Concepción) and smaller
cities. The sample covers participants aged between 18 and 75 years (in
wave 1), representing 77\% of the total national population and 93\% of
the urban areas (ELSOC 2022).

The survey has been conducted yearly since 2016, except for the year
2020, when it was suspended due to the pandemic. The waves 2016, 2017,
2018, 2019, 2022, and 2023 were administered using computer-assisted
personal interviewing (CAPI). However, a reduced version was conducted
using computer-assisted telephone interviews (CATI) in 2021. In
addition, wave 3 included a refreshment sample to counter survey
attrition. As a result, the total sample of wave 3 included 3,748 cases,
of which 2,229 were part of the original sample and 1,519 were from the
refreshment sample. The data from the refreshment sample is not included
in the analytical sample for this study as we analyze longer response
trends. Regarding the original sample, the response rate was 62.4\% in
wave 1, achieving N = 2,927 participants. In broader terms, the
accumulated attrition between wave 1 and wave 7 is 40\%, achieving a
final sample of N = 1,741. The analysis used longitudinal weights to
avoid bias due to the sampling design, allowing us to control for biases
arising from systematic patterns of non-participation in the survey
after the first wave. For a more detailed analysis of responses,
attrition, and the construction of longitudinal weights, visit
https://coes.cl/encuesta-panel/. The dataset is publicly available at:
https://dataverse.harvard.edu/dataverse/elsoc.

\subsection{Variables}\label{variables}

The dependent variable of this study is the justification of educational
inequality. This construct is measured using the following statement:
``It is just that high-income people have a better education for their
children than people with lower incomes'' (``Es justo que las personas
de altos ingresos tengan una mejor educación para sus hijos que las
personas con ingresos más bajos'' in Spanish). Here, the respondents
declared their preferences on a Likert scale from (1) ``strongly
disagree'' to (5) ``strongly agree'' (5). The main independent variables
refer to meritocratic perception, which is operationalized by two items,
one related to effort (``In Chile, people are rewarded for their
efforts''), and the other related to talent (``In Chile, people are
rewarded for their intelligence and skills''). In addition, we included
an indicator for economic inequality perception (``In Chile, the income
differences are too large''). Figure~\ref{fig-frecuencies} shows the
average distribution throughout the years for these variables. We
observe that there is a large group (88.2\%) of respondents who disagree
or strongly disagree that people with higher incomes can access better
education. Likewise, a large group of respondents (91.7\%) agree or
strongly agree that income differences in Chile are too large. On the
other hand, there is an important level of disagreement (disagree +
strongly disagree) about the perception of meritocracy, this is, whether
people are rewarded for their talent (47.6\%) and for their efforts
(57\%).

\begin{figure}[H]

\centering{

\includegraphics[width=0.85\textwidth,height=\textheight]{merit_plot-rr.png}

}

\caption{\label{fig-frecuencies}Frequency of responses for justification
of educational inequality, perception of inequality, and perception of
meritocracy for all the years analyzed}

\end{figure}%

A third group of variables is included in the analysis to assess the
effect of elements related to the pandemic context. These variables were
only included in the 2021 wave, so each individual's responses were
replicated for the remaining years. The first variable addresses the
belief regarding whether, in the context of the pandemic, it is more
important to give priority to economic activity than to the health of
the population, measured on a Likert scale from (1) ``strongly
disagree'' to (5) ``strongly agree''. The second variable is a dummy
(No/Yes) that captures whether individuals received state benefits
during the pandemic (e.g., Food Box, Emergency Family Income,
Contribution for the Middle Class, or others).

For testing the hypothesis regarding status we considered educational
level and household income. Also, we included subjective social status
as it has been argued that perceived social status is a relevant
predictor of attitudes toward economic inequality (Schneider and
Castillo 2015; Castillo et al. 2019). In addition, such subjective
measures can complement objective measures in predicting life chances
(Oesch and Vigna 2023), which are arguably connected to how individuals
experience economic inequality. In this regard, it has been argued that
how people form their views on their standing in society results from
experiences with direct socioeconomic circumstances and social
comparison processes with reference groups (Condon and Wichowsky 2020).
Finally, political identification on a left-right scale, age groups, and
gender were included as control variables. Table~\ref{tbl-descriptives}
shows the whole set of independent variables, the corresponding items,
response categories, and frequencies.

\begin{longtable}[]{@{}l@{}}
\caption{Independent variables}\label{tbl-descriptives}\tabularnewline
\toprule\noalign{}
\endfirsthead
\endhead
\bottomrule\noalign{}
\endlastfoot
\includegraphics{desc02-rr.png} \\
\end{longtable}

\subsection{Methods}\label{methods}

Given the hierarchical structure of the data (observations nested in
survey waves), we estimated a series of longitudinal multilevel models
(Singer and Willett 2009). Such an approach is suited to account for the
shared variance among units (in this case individuals), adjusting the
estimation of the standard errors. The linear multilevel models are
estimated using the R library ``lme4'' (Singer and Willett 2009, 4).

\section{Results}\label{results}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=0.78in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{times}
\usepackage{caption}
\captionsetup[figure, table]{labelfont={bf},labelformat={default},labelsep=period}
\usepackage[figurename=Fig.]{caption}
\usepackage{graphicx}
\usepackage{float}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Changes in the justification of educational inequalities. The role of perceptions of inequality and meritocracy during the COVID pandemic. },
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Changes in the justification of educational inequalities. The
role of perceptions of inequality and meritocracy during the COVID
pandemic. \footnote{Corresponding author
  (\href{mailto:juancastillov@uchile.cl}{\nolinkurl{juancastillov@uchile.cl}}
  -
  \href{https://juancarloscastillo.github.io/jc-castillo/}{jc-castillo.com}
  - Ignacio Carrera Pinto 1045 Ñuñoa, Santiago de Chile). The authors
  recognize the support of the National Research and Development Agency
  (ANID) through the grant from the National Fund for Scientific and
  Technological Development (FONDECYT) N° 1210847 and N° 1250518; from
  the Center for Conflict and Social Cohesion Studies (COES) grant
  Financing Fund for Research Centers in Priority Areas (FONDAP) N°
  1523A0005; and Millenium Nucleus on Digital Inequalities and
  Opportunities (NUDOS) NCS2022\_046}}
\author{}
\date{\vspace{-2.5em}}

\begin{document}
\maketitle

@@
